Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2014-11-14 improved_state_space.tex
edits to model description in intro

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model (henceforth Ricker model), or it might be 
density-independent, such as when growth
is exponential. 
A generic model can be written: 
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= x_0 \nonumber \\ 
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
where $x_0$ is the distribution for initial abundance.  In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian variables: 
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. 
The process variation associated with $N_1$ is often ignored ($x_0 = X_1$), or
the population is assumed to be at stochastic equilibrium such that the $x_0$ can
be regarded as an outcome of the equilibrium distribution \citep{devalpine_hastings:2002}.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
overcome many of the limitations associated with the application of
classical models of population dynamics, 
several problems exist. 
The assumption that random effects are Gaussian distributed makes it
easier to estimate parameters because methods such as the Kalman
filter can be applied 
\citep{dennis_etal:2006}; however, 
these models effectively assume that abundance can be take on any
real value, and hence they can predict negative values of abundance. 
Transforming the count data and the underlying abundance variables 
do little to resolve these issues \citep{ohara_kotze:2010}.

@@

Several extensions of state-space models have been proposed to
overcome the limitations described above. \citet{devalpine_hastings:2002} and
\citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
distributions for the process and observation errors. An observation model 
explicitly modeling
detection probability has been proposed by 
\citet{kery_etal:2009}. \citet{lele_etal:1998} and 
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only
the work by \citet{kery_etal:2009} addressed several of these limitations
simultaneously; however, their model did not include serial
dependence, which is a hallmark of classical population growth models as well as state-space models.

>>>>

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be the observed data. 
%, which differs from $N_t$ due to observation error,
%a random effect denoted $\zeta_t$. 
Temporal variation in $N_t$ is
modeled using a probability distribution whose expected value is
determined by a population growth model, $\mu(N_{t-1}, {\bm \theta})$, such as 
%and random process variation denoted $\eta_t$.
%The growth model may be 
a density-dependent Ricker-logistic model (henceforth Ricker model), or  
a density-independent exponential model with parameter vector $\bm
\theta$. Such a model can be written:  
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
%     N_1 &= x_0 \nonumber \\ 
% N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
% t=2,\hdots,T  \\
% X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
% t=1,\hdots,T. \nonumber 
  N_1 &\sim \mathrm{Normal}(\mu_1, \sigma_0) \nonumber \\
  N_t &\sim \mathrm{Normal}(\mu(N_{t-1}, {\bm \theta}), \sigma) \qquad
  \; \text{for} \;  t=2,\hdots,T  \\
  X_t &\sim \mathrm{Normal}(N_t, \tau) \qquad \qquad \quad \;\;\, \text{for} \;  t=1,\hdots,T  \nonumber
  \end{align}
\end{gather}
\end{linenomath*}
%where $x_0$ is the distribution for initial abundance.  In classical
%state-space models, the two sets of random effects
%are assumed to be i.i.d. Gaussian variables: 
%$\eta_t \sim \mathrm{N}(0, \sigma)$ and
%$\zeta_t \sim \mathrm{N}(0, \tau)$. 
where $\sigma_0$ and $\sigma$ are variance parameters of the
underlying abundance process, and $\tau$ describes variation in 
observation error. 
%Often, parameter estimation requires fixing $\sigma_0$ at zero or assuming
%that initial abundance is an outcome of the theoretical stationary
%distribution of the first-order Markov chain \citep{devalpine_hastings:2002}.  
%The process variation associated with $N_1$ is often ignored ($x_0 = X_1$), or
%the population is assumed to be at stochastic equilibrium such that the $x_0$ can
%be regarded as an outcome of the equilibrium distribution \citep{devalpine_hastings:2002}.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
overcome many of the limitations associated with the application of
classical models of population dynamics, several problems exist. 
The Gaussian assumptions 
% that random effects are Gaussian distributed 
make it easier to estimate parameters because methods such as the Kalman
filter can be applied \citep{dennis_etal:2006}; however, 
these models allow for implausible preditions such as 
% assume that abundance can be take on any real value, and hence they
% can predict 
negative values of abundance. 
Transforming the count data and the underlying abundance variables 
do little to resolve these issues \citep{ohara_kotze:2010}.

@@

Several extensions of state-space models have been proposed to
overcome the limitations described above \citep{newman_etal:2014}. \citet{devalpine_hastings:2002} and
\citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
distributions for the process and observation errors. An observation model 
explicitly modeling
detection probability has been proposed by 
\citet{kery_etal:2009}. \citet{lele_etal:1998} and 
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only
the work by \citet{kery_etal:2009} addressed several of these limitations
simultaneously; however, their model did not include serial
dependence, which is a hallmark of classical population growth models as well as state-space models.
